Live Functional Programming with Typed Holes

### 2018-07-04 abstract.tex
tightened abstract

\emph{Live programming environments} aim to provide programmers (and sometimes audiences) 
with continuous feedback about a program's dynamic behavior as it is being edited. 
%
The problem is that programming languages typically assign dynamic meaning only 
 to programs that are \emph{complete}, i.e. syntactically well-formed and free
of type errors, whereas 
%
during much of the editing process, the program is not yet complete. As a result,  
live feedback flickers in and out or goes stale, 
in many cases for substantial lengths of time.
%
% Defining a semantics for incomplete functional programs would help bridge the
% gap between well-formed states and allow live programming environments to
% provide continuous, uninterrupted, and meaningful feedback.

@@

This paper confronts this problem from type-theoretic first principles by developing a 
\emph{dynamic semantics for incomplete functional programs}, 
based in part on the static semantics for incomplete functional programs
developed in recent work by \citet{popl-paper}. 
%
We model incomplete functional programs as expressions with \emph{holes}, 
with empty holes standing for missing expressions or types, and  non-empty holes 
operating as ``membranes'' around static and dynamic type inconsistencies. 
%
Rather than aborting with an exception when evaluation encounters any of these holes (as in
several existing systems), evaluation proceeds ``around'' the holes,
performing as much of the remaining computation as is possible and tracking the 
closure around each hole instance as it flows through the program. Various editor services can report information from these hole closures 
to help the programmer decide how to fill a hole. 
They also enable a ``fill-and-resume'' feature that 
avoids the need to restart evaluation after edits that amount to hole filling. 
Formally, the semantics draws from both gradual type theory (to handle
type holes) and contextual modal type theory (which provides a
logical foundation for hole closures), and develops additional technical machinery necessary 
to continue evaluation past the various hole forms and to prove important metatheoretic properties.

@@

% , including 
% (1) a type safety property that accounts for expression holes (i.e. free metavariables, 
% from the perspective of CMTT) and cast failure holes (which provide a novel), and (2) a commutativity property that establishes that the
% ``edit-and-resume'' feature is sound.
%
We have mechanized the core formal development using the Agda proof assistant.
%
We also describe a simple implementation, called \HazelnutLive{},
that inserts holes as
necessary during the editing process to guarantee that every edit state has some (possibly incomplete)
type, based in part on the \Hazelnut edit action calculus in prior work by \citet{popl-paper}.
%
Taken together with the type safety property that this paper establishes, the result is 
a proof-of-concept live typed functional programming environment where dynamic feedback 
is truly continuous, i.e. it is available for every possible edit state.
% i.e. \emph{every edit state is guaranteed to have
% non-trivial static and dynamic meaning}.

>>>>

\emph{Live programming environments} aim to provide programmers (and sometimes audiences) 
with continuous feedback about a program's dynamic behavior as it is being edited. 
%
The problem is that programming languages typically assign dynamic meaning only 
 to programs that are \emph{complete}, i.e. syntactically well-formed and free
of type errors. As a result,  
live feedback flickers out or goes stale, 
in many cases for substantial lengths of time.
%
% Defining a semantics for incomplete functional programs would help bridge the
% gap between well-formed states and allow live programming environments to
% provide continuous, uninterrupted, and meaningful feedback.

@@

This paper confronts this problem from type-theoretic first principles by developing 
\emph{a dynamic semantics for incomplete functional programs}, 
based in part on the static semantics for incomplete functional programs
developed in recent work by \citet{popl-paper}. 
%
We model incomplete functional programs as expressions with \emph{holes}, 
with empty holes standing for missing expressions or types, and  non-empty holes 
operating as ``membranes'' around static and dynamic type inconsistencies. 
%
Rather than aborting with an exception when evaluation encounters any of these holes (as in
several existing systems with holes), evaluation proceeds ``around'' the holes,
performing as much of the remaining computation as is possible and tracking the 
closure around each hole instance as it flows through the program. Various editor services can report information from these hole closures 
to help the programmer decide how to fill a hole. 
They also enable a ``fill-and-resume'' feature that 
avoids the need to restart evaluation after edits that amount to hole filling. 
Formally, the semantics draws from both gradual type theory (to handle
type holes) and contextual modal type theory (which provides a
logical basis for hole closures), and develops additional technical machinery necessary 
to continue evaluation past the various hole forms and to prove several important metatheoretic properties.

@@

% , including 
% (1) a type safety property that accounts for expression holes (i.e. free metavariables, 
% from the perspective of CMTT) and cast failure holes (which provide a novel), and (2) a commutativity property that establishes that the
% ``edit-and-resume'' feature is sound.
%
We have mechanized this foundational calculus, called \HazelnutLive{}, using the Agda proof assistant.
%
We also describe a simple implementation 
that inserts holes automatically to guarantee that every editor state has some (possibly incomplete)
type, based on the \Hazelnut edit action calculus of \citet{popl-paper}.
Taken together with the type safety property that we establish, the 
% The 
result is  
a proof-of-concept live programming environment where dynamic feedback 
is truly continuous, i.e. it is available for every possible editor state.
% i.e. \emph{every edit state is guaranteed to have
% non-trivial static and dynamic meaning}.

### 2018-07-04 calculus.tex
tightened abstract

The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$.
%
This form serves to ``box'' an expression of type $\htau_1$ for
treatment as an expression of a consistent type $\htau_2$
(Rule TACast in \Figref{fig:hasType}).%
\footnote{
In the earliest work on gradual type theory, the cast form only gave
the target type~$\htau_2$ \cite{Siek06a}, but it simplifies the dynamic semantics substantially
to include the assigned type~$\htau_1$ in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.
}

>>>>

The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$.
%
This form serves to ``box'' an expression of type $\htau_1$ for
treatment as an expression of a consistent type $\htau_2$
(Rule~\rulename{TACast} in \Figref{fig:hasType}).%
\footnote{
In the earliest work on gradual type theory, the cast form only gave
the target type~$\htau_2$ \cite{Siek06a}, but it simplifies the dynamic semantics substantially
to include the assigned type~$\htau_1$ in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.
}
